Numerical modelling of the Cordilleran ice sheet

### 2014-08-24 kappa.tex
Correct many small mistakes and explain notations.

% Some general low-priority todos
% -- use latex in matplotlib figures
% -- homogeneize subfig label fonts
% -- add software references
% -- redraw PISM interface

@@

\begin{figure*}[b]
  \subgraphics{a}{photo-glacier-crevasses}%
  \hspace{1cm}%
  \subgraphics{b}{photo-glacier-susitna}%
  \hspace{1cm}%
  \subgraphics{c}{photo-glacier-fold}%
  \caption{Illustration of various glacier ice behaviours:
           \textbf{(a)}~surface crevasses formed by ice flow over a hidden
           bedrock knob, Glacier Blanc, French Alps, 2004;
           \textbf{(b)}~medial moraines illustrating downhill ice flow on
           Susitna Glacier, Alaska Range, photo by Austin Post, 1970
           \citep{NSIDC.2009}; and
           \textbf{(c)}~mountain glacier calving into a small lake, revealing a
           folded structure formed by deformed debris layers in the ice,
           Glacier d'Arsine, French Alps, 2004.}
  \label{fig:glacier-mechanics}
\end{figure*}

\begin{figure*}[b]
  \subgraphics{a}{photo-glacial-valley}%
  \hspace{1cm}%
  \subgraphics{b}{photo-erratic-boulder}%
  \hspace{1cm}%
  \subgraphics{c}{photo-melt-channels}%
  \caption{Different indicators of past glaciation:
           \textbf{(a)}~glacial valley, exhibiting a broad, U-shaped profile
           characteristic of erosion by glaciers,
           Cirque du Fer-\`{a}-Cheval, French Alps, 2013;
           \textbf{(b)}~erratic boulders, deposited in an improbable
           configuration as the glacier melted underneath,
           Finse area, Norwegian Scandes, 2011; and
           \textbf{(c)}~lateral meltwater channels, formed on a gentle slope
           as the ice margin retreated towards the valley bottom, Tuya Lake
           area, Canadian Cordillera, 2012.}
  \label{fig:glaciation-indicators}
\end{figure*}

@@

In our daily life, we most commonly experience ice as small, solid cubes
that crackle on contact with refreshing liquids. Yet, on geological
time scales, and under high pressure, ice behaves quite differently.
In the cold and mountainous regions of the globe, ice accumulates through
cumulative snowfall until it reaches a critical thickness. Beyond this
threshold, usually in excess of 50\,m, but dependent on temperature, ice
deforms under its own weight to flow downhill (\cref{fig:glacier-mechanics}).
\emph{Glaciers}, these bodies of moving ice, move by a combination of
meltwater-induced sliding at the base \citep[\S532]{Saussure.1786},
and viscous deformation within the
ice body \citep{Forbes.1846b}.

@@

More recently, palaeoclimate records extracted from deep sea sediments and ice
cores have provided a much more detailed picture of the Earth environmental
history \citep[e.g.,][]{Emiliani.1955, Shackleton.Opdyke.1973,
Dansgaard.etal.1993, Augustin.etal.2004}, indicating neither one, nor four, but
rather tens of glacial cycles. During the last 800000 years (800\,ka), glacial
and interglacial periods have succeeded each other with a 100\,ka periodicity
\citep[\cref{fig:paleo-timeseries};]{Hays.etal.1976, Augustin.etal.2004}.
However, much of the traces left by glaciers on the landscape date from the
most recent of these cycles, the \emph{last glacial cycle}.

@@

Nevertheless, the landform record is sparse in time and, most often, spatially
incomplete. Palaeo-ice sheets did not leave a continuous imprint on the
lanscape, and much of this evidence has been overprinted by subsequent glacier
re-advances and other geomorphological processes \citep{Kleman.1994,
Kleman.etal.2006, Kleman.etal.2010}.
Therefore, geomorphological reconstructions of former glacier extent require a
substantial amount of interpretation. This is where numerical ice sheet models,
incorporating the processes of glacier sliding \citep[e.g.,][]{Weertman.1957}
and deformation \citep[e.g.,][]{Nye.1953} observed on modern glaciers, can
help. Although these models alone contain too many uncertainties to reproduce
accurate ice sheet reconstructions during former glacial cycles
\citep[e.g.,][]{Hebeler.etal.2008}, they can be calibrated against geological
data where they exist \citep{Napieralski.etal.2007}, while allowing
physics-based, quantitative reconstructions of ice sheet sectors where they do
not \citep[e.g.,][]{Marshall.etal.2002, Tarasov.Peltier.2004}.

\begin{figure*}
  \includegraphics{map-northamerica}
  \caption{Relief map of northern North America showing a reconstruction of the
           areas covered by the Cordilleran (CIS), Laurentide (LIS), Innuitian
           (IIS) and Greenland (GIS) ice sheets during the last glacial maximum
           \citep[21.4 to 16.8\,cal\,\chem{^{14}C}\,kyr\,BP,][]{Dyke.2004}.
           The rectangular box delimits the model domain used in all simulation
           and most figures of this thesis. The background
           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
           Data \citep{Patterson.Kelso.2014}.}
  \label{fig:map-northamerica}
\end{figure*}

@@

In this thesis, I focus on numerical modelling of the last Cordilleran ice
sheet, which once covered much of the mountainous terrain of western North
America \citep{Dawson.1888}, a region where topographic complexity has
presented a particular challenge to reconstructing glacial history based on the
geomorphological record \citep[e.g.,][]{Kleman.etal.2010} or numerical ice
sheet modelling \citep{Robert.1991}.

@@

\begin{figure*}
  \subgraphics{a}{photo-vancouver-island}%
  \hspace{1cm}%
  \subgraphics{b}{photo-fraser-valley}%
  \caption{Contrasted climates of the North American Cordillera:
           \textbf{(a)}~Pacific Coast at Vancouver Island, western British
           Columbia, by Martin Margold, 2010.
           \textbf{(b)}~Fraser Valley, south-central British Columbia, 2010.}
  \label{fig:photo-fraser-valley}
\end{figure*}

@@

The \emph{dominant reliefs} occur along two major ribbons that run parallel to
each other along the general orientation of the North American Cordillera. The
western ribbon is formed by the Pacific Coast orogenic belt and includes, from
north to south, the Alaska Range (\cref{fig:photo-alaska-range}), the Wrangell
and Saint-Elias Mountains,
the Coast Mountains and the North Cascades. The eastern ribbon consists of the
Nevadan and Laramide belts, which are separated by the deep and strikingly
linear Rocky Mountain / Tintina Trench depression, and includes the Selwyn and
MacKenzie Mountains in the north, and the Rocky mountains in the south. The
Brooks Range of northern Alaska has been covered by ice, but its glaciers
remained disjunct from the main body of the Cordilleran ice sheet
\citep{Kaufman.Manley.2004}.

The North American Cordillera hosts several \emph{intermontane basins}, the
presence
of which influenced former patterns of Cordilleran ice sheet flow. The two
major depressions are the Fraser Plateau in south-central British Columbia, and
the Liard Lowland in northern British Columbia (\cref{fig:map-cordillera}).
These two basins are separated by the Skeena Mountains, a mildly elevated, but
extensive and deeply incised mountain group. This is the most dominant upland
area in-between the dominant topographic ribbons that form the North American
Cordillera.

@@

The topographic complexity of the North American Cordillera is associated with
distinctive patterns of temperature and precipitation, whose effects on
vegetation are clearly expressed in the landscape
(\cref{fig:photo-fraser-valley}). Due to the large latitudinal extent of the
region, mean annual \emph{temperatures} vary significantly from north to south
(\cref{fig:plot-atm}a). About half of the model domain has mean annual
temperatures below freezing. However, temperature \emph{seasonality},
here computed as the difference between the warmest and the coldest months,
varies between around 10\,\degC\ along most of the Pacific Coast to above
40\,\degC\ for the interior Arctic sectors
(\cref{fig:plot-atm}b). There exists a
partial correlation between cold regions and those that have a large
seasonality, so that many regions where mean annual temperatures are well below
freezing point experience warm summers.

Annual \emph{precipitation rates} are highest on the western slopes of the
Pacific Coast ranges and decline quickly on the eastern side of the mountains
(\cref{fig:photo-fraser-valley,fig:plot-atm}c). In terms of ice sheet
mass balance, this contrast is even reinforced through the differences in
\emph{timing of maximum precipitation} through the year (\cref{fig:plot-atm}d).
Coastal regions receive most precipitation during fall and winter, which is
therefore largely delivered as snow. In contrast, many inland regions
experience dry winters, and most of the annual precipitation, consequently
falls there as rain during summer, further enhancing ice melt.

@@

Evidence from glaciofluvial deposits dated using cosmogenic nuclides at
$2.64^{+0.20}_{-0.18}$\,Ma (uncertainty range from 2.46 to 2.84\,Ma)
indicates the first presence of an ice sheet in the
Cordilleran mountains during the late Pliocene \citep{Hidy.etal.2013}.
This indicates that the North American Cordillera has been glaciated many times
since the late Pliocene, with the first of these glaciations occurring even
before the mid-Pleistocene onset of 100 ka glacial cycles
(\cref{fig:paleo-timeseries}). However, only fragments of the last few
glaciations can be reconstructed from the geologic record, implying that much
of the older evidence has been erased by subsequent glacial re-advances and
perhaps even by other, non-glacial geomorphological processes.

@@

Apart from geological evidence of the final demise of the Cordilleran ice sheet, most field
evidence relates to its LGM extent. It indicates that ice covered most of the
Canadian Cordillera, and some of the adjacent regions in north-western
contiguous USA and Alaska (\cref{fig:map-cordillera}). The Cordilleran ice
sheet originated from the coalescence of several mountain-centred ice caps
\citep{Davis.Mathews.1944}. It is generally thought that this LGM configuration
was short-lived \citep{Clague.etal.1980, Clague.1985, Cosma.etal.2008}, a
hypothesis which can be studied through numerical modelling.

@@

The \emph{western margin} of the Cordilleran ice sheet covered part of the
Pacific continental shelf. An independent ice cap covering Vancouver Island
diverted flow from the main body of ice to drain into the Strait of Georgia
between Vancouver Island and the Olympic Peninsula, where it formed the Juan
de Fuca marine ice lobe, calving into the Pacific Ocean \citep{Herzer.Bornhold.1982}. Further to the
North, the ice margin encroached the Queen Charlotte Islands \citep{Blaise.etal.1990}. However, the
archipelago may not have been fully ice-covered, even during the maximum of
glaciation. Submarine troughs traversing several portions of the continental
shelf ()\cref{fig:map-cordillera}) indicate the erosional action of additional ice streams, yet it is not
clear whether they have formed during the last glacial cycle or during earlier
glaciations.

Finally, the conditions at the \emph{eastern margin} of the Cordilleran ice
sheet are the least understood. Although it is known that the Cordilleran ice
sheet collided with the Laurentide ice sheet at times during the last glacial
cycle \citep[e.g.][]{Margold.etal.2013, Margold.etal.2013a}, the nature of the
interaction between the two ice sheets and the precise location of this
junction remains unclear \citep{Gowan.2013}, while limited information exists
on its timing \citep[e.g.][]{Jackson.etal.1997, Bednarski.Smith.2007}. Because
the present thesis focuses specifically on the Cordilleran ice sheet, I,
therefore, neglect buttressing effects of the Laurentide ice sheet.

The retreat of the ice sheet during the last \emph{deglaciation} was likely
rapid. Most of the evidence for the style and pace of deglaciation comes from
the Fraser Plateau. Studies there have indicated that some of the highest
uplands emerged from the ice sheet prior to the surrounding lowlands, possibly
resulting in thin, stagnant ice remnants covering parts of the plateau
\citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a}. Less
is known about the deglaciation of the northern sector of the Cordilleran ice
sheet, but abundant glacial landforms from the Liard Lowland indicate a
different mode of deglaciation, with active retreat of the ice margin towards
the surrounding mountain ranges \citep{Margold.etal.2013}. Patterns of
deglaciation, as well as the possibility for a late-glacial re-advance, are
discussed in further detail and compared with modelling results in
\CCYC.

@@

This section does not attempt to provide an extensive description of PISM,
which can be found in an on-line documentation
\citep[\url{http://www.pism-docs.org},][]{PISM-authors.2014}.
Instead, it aims to highlight the modelling choices that I have made in this
thesis, and to present an overview of the physics involved in these choices.
Parameter values are summarised in \cref{tab:params}.

\begin{figure}
  \includegraphics[width=80mm]{model-interfaces}
  \caption{PISM is a numerical model for ice and bedrock physics. Processes
           occurring at the interface with the atmosphere and the ocean, that
           are progressive transitions in nature, are parametrised along
           well-defined boundaries in the model.
           From \citet{PISM-authors.2014}.}
  \label{fig:model-interfaces}
\end{figure}

@@

    $q$     & Pseudo-plastic sliding exponent
            & 0.25
            & -
            & - \\

    $v_{th}$& Pseudo-plastic threshold velocity
            & 100.0
            & \unit{m\,yr^{-1}}
            & - \\

    $c_0$   & Till cohesion
            & 0.0
            & Pa
            & - \\

    $\alpha$& Effective pressure coefficient (\CCLI)
            & 0.95
            & -
            & - \\

    $\delta$& Effective pressure coefficient (\CCYC)
            & 0.02
            & -
            & - \\

@@

    $W_{max}$ & Maximal till water thickness
            & 2.0
            & m
            & - \\

@@

    $\nu_m$ & Mantle viscosity
            & 1\e21
            & \unit{Pa\,s}
            & \citet{Lingle.Clark.1985} \\

@@

    $D$     & Lithosphere flexural rigidity
            & 5.0\e24
            & \unit{N}
            & \citet{Lingle.Clark.1985} \\

@@

    $F_s$   & Degree-day factor for ice
            & 3.04\e{-3}
            & \unit{m\,K^{-1}\,day^{-1}}
            & \citet{Shea.etal.2009} \\

@@

    $\gamma$& Air temperature lapse-rate
            & 6\e{-3}
            & \unit{K\,m{-1}}
            & - \\

@@

Thirdly, after defining the strain-rate tensor
${\SRT = \frac{1}{2}(\tgrad{\vv} + \tgrad{\vv}^{\mathrm{T}})}$,
and the deviatoric stress tensor, ${\DST = \CST + p\,\IDT}$,
where $p=-\frac{1}{3}\tr(\CST)$ is the hydrostatic pressure and
$\tens{\delta}$ is the identity tensor, a \emph{constitutive law} for ice
\citep{Nye.1953} based on laboratory experiments \citep{Glen.1952} is used to
relate these two quantities. It reads:
\begin{equation}
    \label{eqn:glenslaw}
    \SRT = A\,\tau_e^{n-1}\,\DST \,.
\end{equation}
The equivalent stress, $\tau_e$, is defined by
${\tau_e}^2 = -\mathrm{II}_{\DST} = \frac{1}{2} \tr(\DST^2)$,
where $\mathrm{II}_{\DST}$ is the second invariant of the stress tensor.
The ice softness coefficient, $A$, depends on ice temperature, $T$, pressure, $p$, and
water content, $\omega$, through an Arrhenius-type law
\citep[Eqs.~63--65]{Paterson.Budd.1982, Aschwanden.etal.2012},
\begin{equation}
    A =
    \begin{cases}
        A_c (1+f\omega)\,e^\frac{-Q_c}{RT_{pa}}
            & \text{if}\ T < T_c \,, \\
        A_w (1+f\omega)\,e^\frac{-Q_w}{RT_{pa}}
            & \text{if}\ T \ge T_c \,,
    \end{cases}
\end{equation}
where $T_{pa}$ is the pressure-adjusted temperature calculated using the
Clapeyron relation, ${T_{pa} = T - \beta p}$. $R$ is the ideal gas constant,
and $A_c$, $A_w$, $Q_c$ and $Q_w$, are constant parameter corresponding to
values measured below and above a critical temperature threshold
${T_c=-10}$\,\degC\ \citep[\cref{tab:params};][]{Paterson.Budd.1982}.
The water fraction, $\omega$, is capped at a maximum value of 0.01, above which
no measurements are available \citep[Eq.~5.7]{Lliboutry.Duval.1985,
Greve.1997}.

@@

Finally, the evolution of temperature within the ice is governed by the
\emph{conservation of energy}. In PISM, the conservation of energy uses an
enthalpy formulation in order to
account for thermodynamic effects associated with internal phase changes in the
glacier \citep[Eqs.~20--21]{Aschwanden.etal.2012}. This enthalpy ($H$)
formulation reads
\begin{equation}
    \label{eqn:enthalpy}
    \rho \left(\matdv{H}\right)
        = -\div{\vect{q}} + \frac{\nu\doteps_e^2}{4} \,,
\end{equation}
where $\vect{q}$ represents the heat flow and
${\nu\dot{\epsilon_e}^2/4 = \tr(\DST\SRT)}$ is a
source term corresponding to strain heat release. In the case where ice
temperature, $T$, is below the pressure-melting point, the heat flow is
modelled by Fourier's law, $\vect{q} = k\,\grad{T}$, where $k$ is the thermal
conductivity of ice, while enthalpy can be expressed as $H=c\,T$, where $c$
denotes the ice specific heat capacity (\cref{tab:params}).
Consequently, the enthalpy equation (\ref{eqn:enthalpy}) can be rewritten using
a more familiar, cold-ice, temperature formulation,
\begin{equation}
    \label{eqn:temperature}
    \matdv{T} = \frac{k}{\rho c} \Delta T
                + \frac{\nu\doteps_e^2}{4\rho c} \,.
\end{equation}
However, note that, here, the term ``enthalpy'' is actually an abuse of
language, commonly used in glaciology as a synonym of ``interval energy''
\citep{Aschwanden.etal.2012}.

@@

% ----------------------------------------------------------------------
\subsection{Simulations of the last glacial cycle}
% ----------------------------------------------------------------------

>>>>

% Some general low-priority todos
% -- use latex in matplotlib figures
% -- homogeneize subfig label fonts
% -- add software references
% -- add Swedish abstract
% -- add numerical implementation

@@

\begin{figure*}[b]
  \subgraphics{a}{photo-glacier-crevasses}%
  \hspace{1cm}%
  \subgraphics{b}{photo-glacier-susitna}%
  \hspace{1cm}%
  \subgraphics{c}{photo-glacier-fold}%
  \caption{Illustration of various glacier ice behaviours:
           \textbf{(a)}~surface crevasses formed by ice flow over a hidden
           bedrock knob, Glacier Blanc, French Alps, 2004;
           \textbf{(b)}~medial moraines illustrating downhill ice flow on
           Susitna Glacier, Alaska Range, photograph by Austin Post, 1970
           \citep{NSIDC.2009}; and
           \textbf{(c)}~mountain glacier calving into a small lake, revealing a
           folded structure formed by deformed debris layers in the ice,
           Glacier d'Arsine, French Alps, 2004.}
  \label{fig:glacier-mechanics}
\end{figure*}

\begin{figure*}[b]
  \subgraphics{a}{photo-glacial-valley}%
  \hspace{1cm}%
  \subgraphics{b}{photo-erratic-boulder}%
  \hspace{1cm}%
  \subgraphics{c}{photo-melt-channels}%
  \caption{Different indicators of past glaciation:
           \textbf{(a)}~glacial valley, exhibiting a broad, U-shaped profile
           characteristic of erosion by glaciers,
           Cirque du Fer-\`{a}-Cheval, French Alps, 2013;
           \textbf{(b)}~erratic boulders, deposited in an improbable
           configuration as the glacier melted underneath,
           Finse area, Norwegian Scandes, 2011; and
           \textbf{(c)}~lateral meltwater channels, formed on a gentle slope
           as the lateral ice margin retreated towards the valley bottom, Tuya
           Lake area, Canadian Cordillera, 2012.}
  \label{fig:glaciation-indicators}
\end{figure*}

@@

In our daily life, we most commonly experience ice as small, solid cubes
that crackle on contact with refreshing liquids. Yet, on geological
time scales, and under high pressure, ice behaves quite differently.
In the cold and mountainous regions of the globe, ice accumulates through
cumulative snowfall until it reaches a critical thickness. Beyond this
threshold, usually of several tens of metres, but dependent on temperature, ice
deforms under its own weight to flow downhill (\cref{fig:glacier-mechanics}).
\emph{Glaciers}, these bodies of moving ice, flow by a combination of
meltwater-induced sliding at the base \citep[\S532]{Saussure.1786},
and viscous deformation within the
ice body \citep{Forbes.1846b}.

@@

More recently, palaeoclimate records extracted from deep sea sediments and ice
cores have provided a much more detailed picture of the Earth environmental
history \citep[e.g.,][]{Emiliani.1955, Shackleton.Opdyke.1973,
Dansgaard.etal.1993, Augustin.etal.2004}, indicating neither one, nor four, but
rather tens of glacial cycles. During the last 800000 years (800\,ka), glacial
and interglacial periods have succeeded each other with a 100\,ka periodicity
\citep[\cref{fig:paleo-timeseries};][]{Hays.etal.1976, Augustin.etal.2004}.
However, much of the traces left by glaciers on the landscape date from the
most recent of these cycles, the \emph{last glacial cycle}.

@@

Nevertheless, the landform record is sparse in time and, most often, spatially
incomplete. Palaeo-ice sheets did not leave a continuous imprint on the
lanscape, and much of this evidence has been overprinted by subsequent glacier
re-advances and other geomorphological processes \citep{Kleman.1994,
Kleman.etal.2006, Kleman.etal.2010}.
Therefore, geomorphological reconstructions of former glacier extent require a
substantial amount of interpretation. This is where numerical ice sheet models,
incorporating the processes of glacier sliding \citep[e.g.,][]{Weertman.1957}
and deformation \citep[e.g.,][]{Nye.1953} observed on modern glaciers, can
help. Although these models alone contain too many uncertainties to reproduce
accurate ice sheet reconstructions during former glacial cycles
\citep[e.g.,][]{Hebeler.etal.2008}, they can be calibrated against geological
data where they exist \citep{Napieralski.etal.2007}, while allowing
quantitative reconstructions of ice sheet sectors where they do not, based on
approximated physics of glacier flow \citep[e.g.,][]{Marshall.etal.2002,
Tarasov.Peltier.2004}.

\begin{figure*}
  \includegraphics{map-northamerica}
  \caption{Relief map of northern North America showing a reconstruction of the
           areas covered by the Cordilleran (CIS), Laurentide (LIS), Innuitian
           (IIS) and Greenland (GIS) ice sheets during the last glacial maximum
           \citep[21.4 to 16.8\,cal\,\chem{^{14}C}\,kyr\,BP,][]{Dyke.2004}.
           The rectangular box delimits the model domain used in all
           simulations and most figures of this thesis. The background
           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
           Data \citep{Patterson.Kelso.2014}.}
  \label{fig:map-northamerica}
\end{figure*}

@@

In this thesis, I focus on numerical modelling of the last Cordilleran ice
sheet, which once covered much of the mountainous terrain of western North
America \citep{Dawson.1888}, a region where topographic complexity has
presented a particular challenge to reconstructing glacial history based on the
geomorphological record \citep[e.g.,][]{Kleman.etal.2010} or numerical ice
sheet modelling \citep[e.g.,][]{Robert.1991}.

@@

\begin{figure*}
  \subgraphics{a}{photo-vancouver-island}%
  \hspace{1cm}%
  \subgraphics{b}{photo-fraser-valley}%
  \caption{Contrasted climates of the North American Cordillera:
           \textbf{(a)}~Pacific Coast at Vancouver Island, western British
           Columbia, photograph by Martin Margold, 2010.
           \textbf{(b)}~Fraser Valley, south-central British Columbia, 2010.}
  \label{fig:photo-fraser-valley}
\end{figure*}

@@

The \emph{dominant reliefs} occur along two major ribbons that run parallel to
each other along the general orientation of the North American Cordillera. The
western ribbon is formed by the Pacific Coast orogenic belt and includes, from
north to south, the Alaska Range (\cref{fig:photo-alaska-range}), the Wrangell
and Saint-Elias Mountains,
the Coast Mountains and the North Cascades. The eastern ribbon consists of the
Nevadan and Laramide belts, which are separated by the deep and strikingly
linear Rocky Mountain~/ Tintina Trench depression, and includes the Selwyn and
MacKenzie Mountains in the north, and the Rocky mountains in the south. The
Brooks Range of northern Alaska has been covered by ice, but its glaciers
remained disjunct from the main body of the Cordilleran ice sheet
\citep{Kaufman.Manley.2004}.

The North American Cordillera hosts several \emph{intermontane basins}, the
presence
of which influenced former patterns of Cordilleran ice sheet flow. The two
major depressions are the Interior Plateau in south-central British Columbia,
and the Liard Lowland in northern British Columbia (\cref{fig:map-cordillera}).
These two basins are separated by the Skeena Mountains, a mildly elevated, but
extensive and deeply incised mountain group. This is the most dominant upland
area in-between the dominant topographic ribbons that form the North American
Cordillera.

@@

The topographic complexity of the North American Cordillera is associated with
distinctive patterns of temperature and precipitation, whose effects on
vegetation are clearly expressed in the landscape
(\cref{fig:photo-fraser-valley}). Due to the large latitudinal extent of the
region, mean annual \emph{temperatures} vary significantly from north to south
(\cref{fig:plot-atm}a). About half of the model domain has mean annual
temperatures below freezing. However, temperature \emph{seasonality},
here computed as the difference between the warmest and the coldest months,
varies between around 10\,\degC\ along most of the Pacific Coast to above
40\,\degC\ for the continental Arctic sectors
(\cref{fig:plot-atm}b). There exists a
partial correlation between cold regions and those that have a large
seasonality, so that many regions where mean annual temperatures are well below
freezing point experience warm summers.

Annual \emph{precipitation rates} are highest on the western slopes of the
Pacific Coast ranges and decline quickly on the eastern side of the mountains
(\cref{fig:photo-fraser-valley,fig:plot-atm}c). In terms of ice sheet
mass balance, this contrast is even reinforced through the differences in
\emph{timing of maximum precipitation} through the year (\cref{fig:plot-atm}d).
Coastal regions receive most precipitation during fall and winter, which is
therefore largely delivered as snow. In contrast, many inland regions
experience dry winters, and most of the annual precipitation consequently
falls there as rain during summer, further enhancing ice melt.

@@

Evidence from glaciofluvial deposits dated using cosmogenic nuclides at
$2.64^{+0.20}_{-0.18}$\,Ma (uncertainty range from 2.46 to 2.84\,Ma)
indicates the first presence of an ice sheet on the
Cordilleran mountains during the late Pliocene \citep{Hidy.etal.2013}.
This indicates that the North American Cordillera has been glaciated many times
since the late Pliocene, with the first of these glaciations occurring even
before the mid-Pleistocene onset of 100\,ka glacial cycles
(\cref{fig:paleo-timeseries}). However, only fragments of the last few
glaciations can be reconstructed from the geologic record, implying that much
of the older evidence has been erased by subsequent glacial re-advances and
perhaps even by other, non-glacial geomorphological processes.

@@

Apart from geological evidence of the final demise of the Cordilleran ice sheet, most field
evidence relates to its LGM extent. It indicates that ice covered most of the
Canadian Cordillera, and some of the adjacent regions in north-western
contiguous USA and Alaska (\cref{fig:map-cordillera}). The LGM Cordilleran ice
sheet originated from the coalescence of several mountain-centred ice caps
\citep{Davis.Mathews.1944}. It is generally thought that the ice sheet
configuration
was short-lived \citep{Clague.etal.1980, Clague.1985, Cosma.etal.2008}, a
hypothesis which can be studied through numerical modelling.

@@

The \emph{western margin} of the Cordilleran ice sheet covered part of the
Pacific continental shelf. An independent ice cap covering Vancouver Island
diverted flow from the main body of ice to drain into the Strait of Georgia and
then between Vancouver Island and the Olympic Peninsula, where it formed the
Juan de Fuca marine ice lobe, calving into the Pacific Ocean
\citep{Herzer.Bornhold.1982}. Further to the North, the ice margin encroached
the Queen Charlotte Islands \citep{Blaise.etal.1990}. However, the
archipelago may not have been fully ice-covered, even during the maximum of
glaciation. Submarine troughs traversing several portions of the continental
shelf (\cref{fig:map-cordillera}) indicate the erosional action of additional
ice streams, yet it is not
clear whether they have formed during the last glacial cycle or during earlier
glaciations.

Finally, the conditions at the \emph{eastern margin} of the Cordilleran ice
sheet are the least understood. Although it is known that the Cordilleran ice
sheet collided with the Laurentide ice sheet at times during the last glacial
cycle \citep[e.g.][]{Margold.etal.2013, Margold.etal.2013a}, the nature of the
interaction between the two ice sheets and the precise location of this
junction remains unclear \citep{Gowan.2013}, while limited information exists
on its timing \citep[e.g.,][]{Jackson.etal.1997, Bednarski.Smith.2007}. Because
the present thesis focuses specifically on the Cordilleran ice sheet, I,
therefore, neglect buttressing effects of the Laurentide ice sheet.

The retreat of the ice sheet during the last \emph{deglaciation} was likely
rapid. Most of the evidence for the style and pace of deglaciation comes from
the Interior Plateau of British Columbia. Studies there have indicated that
some of the highest
uplands emerged from the ice sheet prior to the surrounding lowlands, possibly
resulting in thin, stagnant ice remnants covering parts of the plateau
\citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a}. Less
is known about the deglaciation of the northern sector of the Cordilleran ice
sheet, but abundant glacial landforms from the Liard Lowland indicate a
different mode of deglaciation, with active retreat of the ice margin towards
the surrounding mountain ranges \citep{Margold.etal.2013}. Patterns of
deglaciation, as well as the possibility for a late-glacial re-advance, are
discussed in further detail and compared with modelling results in
\CCYC.

@@

This section does not attempt to provide an extensive description of PISM,
which can be found in an on-line documentation
\citep[\url{http://www.pism-docs.org},][]{PISM-authors.2014}.
Instead, it aims to highlight the modelling choices that I have made in this
thesis, and to present an overview of the physics involved in these choices.

Vectors are identified by bold arrow symbols ($\vv$) and second-order tensors by bold symbols ($\CST$), while scalars use a normal font ($s$). The divergence of a vector field is written $\div{\vv}$, and $\grad{s}$ denotes the gradient of a scalar field. The divergence of a tensor field is characterised by $\tdiv{\CST}$ and the gradient of a vector field by $\tgrad{\vv}$. The race of a tensor $\CST$ is written $\tr(\CST)$. Parameter values are summarised in \cref{tab:params}.

\newpage  %FIXME temporary fix

\begin{figure}
  \includegraphics[width=80mm]{model-interfaces}
  \caption{PISM is a numerical model for ice and bedrock physics. Processes
           occurring at the interface with the atmosphere and the ocean, that
           are progressive transitions in nature, are parametrised along
           well-defined boundaries in the model.
           After \citet{PISM-authors.2014}.}
  \label{fig:model-interfaces}
\end{figure}

@@

    $q$     & Pseudo-plastic sliding exponent
            & 0.25
            & -
            & \citet{Aschwanden.etal.2013} \\

    $v_{th}$& Pseudo-plastic threshold velocity
            & 100.0
            & \unit{m\,yr^{-1}}
            & \citet{Aschwanden.etal.2013} \\

    $c_0$   & Till cohesion
            & 0.0
            & Pa
            & \citet{Aschwanden.etal.2013} \\

    $\alpha$& Effective pressure coefficient (\CCLI)
            & 0.95
            & -
            & \citet{Aschwanden.etal.2013} \\

    $\delta$& Effective pressure coefficient (\CCYC)
            & 0.02
            & -
            & \citet{Bueler.Pelt.2014} \\

@@

    $W_{max}$ & Maximal till water thickness
            & 2.0
            & m
            & \citet{Aschwanden.etal.2013} \\

@@

    $\nu_m$ & Mantle viscosity
            & 1\e{21}
            & \unit{Pa\,s}
            & \citet{Lingle.Clark.1985} \\

@@

    $D$     & Lithosphere flexural rigidity
            & 5.0\e{24}
            & \unit{N}
            & \citet{Lingle.Clark.1985} \\

@@

    $F_s$   & Degree-day factor for snow
            & 3.04\e{-3}
            & \unit{m\,K^{-1}\,day^{-1}}
            & \citet{Shea.etal.2009} \\

@@

    $\gamma$& Air temperature lapse rate
            & 6\e{-3}
            & \unit{K\,m{-1}}
            & - \\

@@

Thirdly, after defining the strain-rate tensor
${\SRT = \frac{1}{2}(\tgrad{\vv} + \tgrad{\vv}^{\mathrm{T}})}$,
and the deviatoric stress tensor, ${\DST = \CST + p\,\IDT}$,
where $p=-\frac{1}{3}\tr(\CST)$ is the hydrostatic pressure and
$\tens{\delta}$ is the identity tensor, a \emph{constitutive law} for ice
\citep{Nye.1953} based on laboratory experiments \citep{Glen.1952} is used to
relate these two quantities. It reads:
\begin{equation}
    \label{eqn:glenslaw}
    \SRT = A\,\tau_e^{n-1}\,\DST \,.
\end{equation}
The equivalent stress, $\tau_e$, is defined by
${\tau_e}^2 = -\mathrm{II}_{\DST} = \frac{1}{2} \tr(\DST^2)$,
where $\mathrm{II}_{\DST}$ is the second invariant of the stress tensor.
The ice softness coefficient, $A$, depends on ice temperature, $T$, pressure, $p$, and
water content, $\omega$, through an Arrhenius-type law
\citep[Eqs.~63--65]{Paterson.Budd.1982, Aschwanden.etal.2012},
\begin{equation}
    A =
    \begin{cases}
        A_c (1+f\omega)\,e^\frac{-Q_c}{RT_{pa}}
            & \text{if}\ T < T_c \,, \\
        A_w (1+f\omega)\,e^\frac{-Q_w}{RT_{pa}}
            & \text{if}\ T \ge T_c \,,
    \end{cases}
\end{equation}
where $T_{pa}$ is the pressure-adjusted temperature calculated using the
Clapeyron relation, ${T_{pa} = T - \beta p}$. $R$ is the ideal gas constant,
and $A_c$, $A_w$, $Q_c$ and $Q_w$, are constant parameters corresponding to
values measured below and above a critical temperature threshold
${T_c=-10}$\,\degC\ \citep[\cref{tab:params};][]{Paterson.Budd.1982}.
The water fraction, $\omega$, is capped at a maximum value of 0.01, above which
no measurements are available \citep[Eq.~5.7]{Lliboutry.Duval.1985,
Greve.1997}.

@@

Finally, the evolution of temperature within the ice is governed by the
\emph{conservation of energy}. In PISM, the conservation of energy uses an
enthalpy formulation in order to
account for thermodynamic effects associated with internal phase changes in the
glacier \citep[Eqs.~20--21]{Aschwanden.etal.2012}. This enthalpy ($H$)
formulation reads
\begin{equation}
    \label{eqn:enthalpy}
    \rho \left(\matdv{H}\right)
        = -\div{\vect{q}} + \frac{\nu\doteps_e^2}{4} \,,
\end{equation}
where $\vect{q}$ represents the heat flow and
${\nu\dot{\epsilon_e}^2/4 = \tr(\DST\SRT)}$ is a
source term corresponding to strain heat release. In the case where ice
temperature, $T$, is below the pressure-melting point, the heat flow is
modelled by Fourier's law, $\vect{q} = k\,\grad{T}$, where $k$ is the thermal
conductivity of ice, while enthalpy can be expressed as $H=c\,T$, where $c$
denotes the ice specific heat capacity (\cref{tab:params}).
Consequently, the enthalpy equation (\ref{eqn:enthalpy}) can be rewritten using
a perhaps more familiar, cold-ice, temperature formulation,
\begin{equation}
    \label{eqn:temperature}
    \matdv{T} = \frac{k}{\rho c} \Delta T
                + \frac{\nu\doteps_e^2}{4\rho c} \,.
\end{equation}
Note that, here, the term ``enthalpy'' is actually an abuse of
language, commonly used in glaciology as a synonym of ``interval energy''
\citep{Aschwanden.etal.2012}.

@@

% ----------------------------------------------------------------------
\subsection{Simulations of the last glacial cycle}
\label{sec:ccyc}
% ----------------------------------------------------------------------
